---
title: "Factors In-Class Activity"
author: "Eric Bouma"
format: html
editor: visual
---

Using the forcats library, use your Rice Center Data function to add these changes.

Change the order of the weekdays to start with Monday.
Add two additional factor columns:
Weekday/Weekend based on the day of the week.
Create a column based on the *four primary lunar phases* for the data collected.
[Superstar Question] Create a column that indicates whether the tide is coming in (Incoming) or going out (Outgoing).



```{r, echo = FALSE, include = FALSE}

#Bringing in data set using our function

source("getRiceData.R")
rice <- getRiceData()
```


Change the order of the weekdays to start with Monday.
```{r}
rice |>
mutate(Weekday = fct_recode(Weekday,
                               "Monday" = "Sunday",
                               "Tuesday" = "Monday",
                               "Wednesday" = "Tuesday",
                               "Thursday" = "Wednesday",
                               "Friday" = "Thursday",
                               "Saturday" = "Friday",
                               "Sunday" = "Saturday")) |>
  mutate(Weekday = fct_relevel(Weekday,
                                "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")) -> rice

levels(rice$Weekday)
```

Weekday/Weekend based on the day of the week.
```{r}
rice |>
  mutate( `Weekend/Weekday` = fct_lump(Weekday,
                                     n = 2,
                                     other = "weekend")) -> rice
    rice$`Weekend/Weekday` <- fct_recode( rice$`Weekend/Weekday`,
                           "weekend" = "Sunday",
                           "weekend" = "Saturday",
                           "weekday" = "Monday",
                           "weekday" = "Tuesday",
                           "weekday" = "Wednesday",
                           "weekday" = "Thursday",
                           "weekday" = "Friday")
```

Create a column based on the *four primary lunar phases* for the data collected
```{r}
#lunar package
```

Create a column that indicates whether the tide is coming in (Incoming) or going out (Outgoing)
```{r}

```

