---
title: "Final Exam Written Document"
author: "Eric Bouma"
format: html
date: 12/16/24
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library( tidyverse )
library( lubridate )
library( dplyr )
library( knitr )
library( kableExtra )
library(ggspatial)
library(prettymapr)

```

```{r, echo=FALSE, message=FALSE}

read_csv("Arapat_Locations.csv") -> ara.locations
read_csv("Arapat_Samples.csv") -> ara.samples

```

## Abstract

The Sonora Desert bark beetle, *Araptus attenuatus*, is a tiny beetle that is found exclusively in the Sonora desert in parts of Mexico and the United States. This beetle has an interesting life cycle, as it lives inside a specific species of plant, *Euphoria lomelli*, and no where else. The males will hollow out a chamber inside of newly senescing stems of *E. lomelli* to live in. Once they have a home, males then produce a hormone that attracts females for reproduction. Both male and female beetles can fly, however, once a male gets a home established, they seem to stay in one place and it is the females that move around more. The overall motivation for this report is to determine the extent to which changing climate may have impacted sex ratio divergence of *Araptus attenuatus* among locations in the Sonora Desert.

## Methods

The data sets used include a .CSV file (`Ara Locations`) of point locations for all 31 sampling sites, a .CSV (`Ara Samples`) of the raw characteristic beetle data for all 31 sites, a GeoTIFF (`Suitability now`) raster of estimated habitat suitability index, and a .ASC (`Suitability lmg`) raster of historical estimates for habitat suitability during the last glacial maximum.   
To help determine if a changing climate influences the sex ratio, a couple of questions need to be answered:

1) Do sampled populations of the beetle have different sex ratios? The `tidyverse` suite of functions was used to group the "Ara samples" data by each site and then calculate a new column for the male to female ratio. Specifically, the `mutate`, `summarise`, and `group by` functions. `ggplot` was used to display the results graphically via a bar chart. 
2) Which sites have sex ratios that deviate from equal proportions of males and females at the site? The `ggplot` graph from the previous question was used to show the deviation from a perfect 1:1 ratio. A red line was applied at 1 (perfect ratio) to show which sites had an equal number of males to females. 
3) Is there a functional relationship between the habitat suitability at the sampling locations and the sex ratio? The `Ara locations` data set was used to make a `sf` object called `sites`. Then the "`Suitability now` raster was read in, and the `extract` function was used to give each `sf` point in `sites` an attributed suitability value from the raster. `QQnorm` and `ggplot` graphs were used to visualize the data to determine the statistics to be ran. A `lm` was used to see if there was a relationship between `MFRatio ~ HabitatSuitPresent`. Once the model was created, the $R^2$ and $p$-value were calculated to determine if the model was "strong" or not. A `kable` table was used to display the results.   
4) Does the inclusion of Phenotype A and Phenotype B improve the functional relationship over habitat suitability alone? `Tidyverse` was used again on the `Ara samples` data to create two new columns, `Phenotype A` and `Phenotype B` for each site using the `summarise` and `groupby` functions. Data was visualized using `QQnorm` and `histograms` to check for normality and residual behavior. Four `lm` models were created; `MFRatio ~ Phenotype A`, `MFRatio ~ Phenotype B`, `MFRatio ~ Phenotype A` + `Phenotype B`, and `MFRatio ~ Phenotype A` + `Phenotype B` + `Habitat Suitability`. These four models were then put into a `data.frame` for further analysis where the $R^2$, $p$-value, $AIC$, and $ΔAIC$ were calculated for each. Finally, the data was displayed in a `kable` table.  
5) Has the suitability changed from 20,000 years ago? The `Suitability lmg` raster was read in, and the `extract` function was used again to retrieve historical habitat suitability scores. The `sites` data set was used again to attribute the values to the 31 different sites. `sites`  has the present day habitat scores already from question 3, and a `ggplot` bar graph was used to visualize the difference. Red bars represent a decrease in habitat score, and a green bar represent an increase in habitat score.  
6) What was the historical sex ratio, and are there trends due to climatic changes? The `sites` data set was merged with the data set from question one containing `MFRatio`. `Mutate` was then used to calculate two new columns named `ΔHabitat` and `ΔMFRatio`. `ΔHabitat` was calculated by subtracting the historical value from the present day habitat value. `ΔMFRatio` was calculated by first using a `lm` to determine the current relationship between `MFRatio ~ HabitatSuitPresent + HabitatSuitHistorical`. Then the `predict` function was used to determine the historical `MFRatio`. After the historical was predicted, then it was subtracted from the present day to get the change. `QQnorm`, `histogram`, and `shapiro.test` were used to visualize the behavior of the data and residuals to determine the correct statistical test. A `spearmans cor.test` was used to determine the significance of the relationship between `ΔHabitat` and `ΔMFRatio`. Finally, multiple `ggplot` graphs were used to visualize the differences in the historical vs present day `Habitat Suitability` and the `MFRatio` through a map overlay. 

## Results

Q1: The different sampled populations (different sites) do have different sex ratios as demonstrated by the `ggplot` below (Figure 1). 
```{r, echo=FALSE}

ara.samples |>
  group_by(Site) |>
  summarise( Males = sum(Sex == "Male"),
             Females = sum(Sex == "Female")) |>
  mutate( MFRatio = Males/Females)-> Q1

Q1$SiteNum <- as.numeric(gsub("Site", "", Q1$Site))

Q1 |>
  arrange( SiteNum ) -> Q1

```

```{r, echo=FALSE}

Q1 |>
    ggplot( aes(x = factor(SiteNum), y = MFRatio)) + 
    labs ( title = "Male to Female Ratio at Each Site",
           x = "Site",
           y = "Ratio") + 
    theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
    geom_col(col = "black",
            fill = "blue") +
  scale_x_discrete(labels = as.character(1:31))

```
Figure 1. Male to female ratio at each site.

Q2: Of the 31 sites, 29 sites had sex ratios that deviate from equal proportions of males to females. Sites 23 and 25 are the only two which have an equal 1:1 proportion (Figure 2).
```{r, echo=FALSE, warning=FALSE}

Q1 |>
    ggplot( aes(x = factor(SiteNum), y = MFRatio)) + 
    labs ( title = "Male to Female Ratio at Each Site",
           x = "Site",
           y = "Ratio") + 
    theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
    geom_col(col = "black",
            fill = "blue") +
  scale_x_discrete(labels = as.character(1:31)) +
  geom_hline(yintercept = 1, color = "red", linetype = "solid", size = 0.5)


```
Figure 2. Male to female ratio at each site.

Q3: The `QQnorm` and `histogram` plots tell us that the data is fairly well behaved, with the residuals being fairly tight along the dotted line, and the histograms being decently bell-curved, both indicating normal distributed. This informs us that our resulting $p$-values are valid. 
```{r, echo=FALSE, message=FALSE}

library(raster)
library(sf)

#read in raster
r <- raster("suitability_now.tif")
#proj raster
new.proj <- "+proj=longlat +datum=WGS84 +no_defs +type=crs"

r.utm <- projectRaster(r, crs = new.proj)
#points as sf and correct crs
ara.locations |>
  st_as_sf( coords=c("Longitude","Latitude"), 
          crs=4326 ) -> sites
#pull out raster data for each site
extract( r.utm, sites) -> habitat_score

sites$HabitatSuitPresent <- habitat_score

merge( sites, Q1, by = "Site", all = FALSE) -> Q3

```

```{r, echo=FALSE}
#p-value function
get_pval <- function( model ) {
  f <- summary( model )$fstatistic[1]
  df1 <- summary( model )$fstatistic[2]
  df2 <- summary( model )$fstatistic[3]
  p <- as.numeric( 1.0 - pf( f, df1, df2 ) )
  return( p  )
}

```

```{r, echo=FALSE}
fit.ratio <- lm(MFRatio ~ HabitatSuitPresent, data = Q3)

par(mfrow = c(1,3))
plot(fit.ratio, which = 2, main = "MFRatio ~ HabitatSuitPresent")
hist(Q3$MFRatio, main = "Male to Female Ratio")
hist(Q3$HabitatSuitPresent, main = "Present Day Habitat Suitability")
par(mfrow = c(1,1))

```
Figure 3. `QQnorm` plot showing the `lm` residual distribution, and `histograms` for both individual variables. 
```{r, echo=FALSE}
data.frame( Model = c("Male to Female Ratio ~ Habitat Suitability"),
            R2 = c(summary(fit.ratio)$r.squared),
            P = c( get_pval( fit.ratio) )) -> df.models.Q3
```

The `kable` table for the `lm` of `MFRatio ~ Habitat Suitability` tells us that although only `r round(df.models.Q3$R2 *100, 1)`% of the variation of `MFRatio ` is explained by `HabitatSuitPresent`, there is a significant relationship between the two variables with a $p$ = `r round(df.models.Q3$P,4)`.
```{r, echo=FALSE}
data.frame( Model = c("Male to Female Ratio ~ Habitat Suitability"),
            R2 = c(summary(fit.ratio)$r.squared),
            P = c( get_pval( fit.ratio) )) -> df.models.Q3

df.models.Q3 |>
  arrange( -R2 ) |>
  mutate( P = format( P, scientific=TRUE, digits=3)) |>
  kable( caption = "<b>Male to Female Ratio Explained by Habitat Suitability<b>",
         digits = 3) |>
   kable_classic( full_width = FALSE,
                 lightable_options = "striped") |>
  row_spec( 0, bold = TRUE)
```
Table 1. Male to Female Ratio Explained by Habitat Suitability.

Q4: The `QQnorm` and `histogram` plots tell us that `Phenotype A` and `Phenotype B`are well behaved as individual variables. The residuals are decently tight along the dotted line, and the histograms being both normally distributed. The addition of multiple variables to the equations still shows residuals that are normal for both `Phenotype A + Phenotype B`, and `Phenotype A + Phenotype B + Habitat Suitability`. Again, the normal residuals informs us that our resulting $p$-value is valid. 
```{r, echo=FALSE}
#data wrangle 
ara.samples |>
  group_by(Site) |>
  summarise( PhenotypeA = mean(PhenotypeA),
             PhenotypeB = mean(PhenotypeB)) -> Q4

merge( Q3,Q4, by = "Site", all = FALSE) -> Q4.1

Q4.1 |>
  arrange( SiteNum ) -> Q4.1

#new models
fit.PhenoA <- lm(MFRatio ~ PhenotypeA, data = Q4.1)
fit.PhenoB <- lm(MFRatio ~ PhenotypeB, data = Q4.1)
fit.PhenoBoth <- lm(MFRatio ~ PhenotypeA + PhenotypeB, data = Q4.1)
fit.All <- lm(MFRatio ~ PhenotypeA + PhenotypeB + HabitatSuitPresent, data = Q4.1)

# data distribution visualization
par(mfrow = c(1,2))
plot(fit.PhenoA, which = 2, main = "Phenotype A")
hist(Q4.1$PhenotypeA, main = "Phenotype A")
plot(fit.PhenoB, which = 2, main = "Phenotype B")
hist(Q4.1$PhenotypeB, main = "Phenotype B")
plot(fit.PhenoBoth, which = 2, main = "Phenotype A + B")
plot(fit.All, which = 2, main = "A + B + Habitat")
par(mfrow = c(1,1))


```
Figure 4. `QQnorm` plot showing residual distribution, and histograms for individual variable data behavior.
```{r, echo=FALSE}

data.frame( Model = c("Male to Female Ratio ~ Habitat Suitability",
                      "Male to Female Ratio ~ Phenotype A",
                      "Male to Female Ratio ~ Phenotype B",
                      "Male to Female Ratio ~ Phenotype A + Phenotype B",
                      "Male to Female Ratio ~ Phenotype A + Phenotype B + Habitat Suitability"),
            R2 = c(summary(fit.ratio)$r.squared,
                   summary(fit.PhenoA)$r.squared,
                   summary(fit.PhenoB)$r.squared,
                   summary(fit.PhenoBoth)$r.squared,
                   summary(fit.All)$r.squared),
            P = c( get_pval( fit.ratio),
                   get_pval( fit.PhenoA),
                   get_pval( fit.PhenoB),
                   get_pval( fit.PhenoBoth),
                   get_pval( fit.All) )) -> df.models.Q4

#####AIC Calc######################################################################################

data.frame( Model2 = c( "fit.ratio",
     "fit.PhenoA",
     "fit.PhenoB",
     "fit.PhenoBoth",
     "fit.All"),
     model = I(list(fit.ratio,
                    fit.PhenoA,
                    fit.PhenoB,
                    fit.PhenoBoth,
                    fit.All))) -> models.aic

#R-sq and P-val calc
models.aic |>
  mutate(Model = c( "Male to Female Ratio ~ Habitat Suitability",
                      "Male to Female Ratio ~ Phenotype A",
                      "Male to Female Ratio ~ Phenotype B",
                      "Male to Female Ratio ~ Phenotype A + Phenotype B",
                      "Male to Female Ratio ~ Phenotype A + Phenotype B + Habitat Suitability"))|>
  mutate (R2 = c( summary(fit.ratio)$r.squared,
                   summary(fit.PhenoA)$r.squared,
                   summary(fit.PhenoB)$r.squared,
                   summary(fit.PhenoBoth)$r.squared,
                   summary(fit.All)$r.squared)) |>
  mutate(  P = c( get_pval( fit.ratio),
                   get_pval( fit.PhenoA),
                   get_pval( fit.PhenoB),
                   get_pval( fit.PhenoBoth),
                   get_pval( fit.All ))) -> models.aic
          
#Calc AIC
models.aic$AIC <- sapply(models.aic$model, AIC)

#Calc delta AIC
min_AIC <- min(models.aic$AIC)
models.aic$ΔAIC <- models.aic$AIC - min_AIC 

```

The `kable` table below tells us that the variable with the best functional relationship is `Habitat Suitability` (Table 2). `Habitat Suitability` has the only significant $p$-value, $p$ = `r round(models.aic[1,"P"],4)`, (I.E. there is a relationship between `MFRatio` and `Habitat Suitability`) and it has the lowest $AIC$, $AIC$ = `r round(models.aic[1,"AIC"],3)`, and $ΔAIC$, $ΔAIC$ = `r models.aic[1,"ΔAIC"]`. The lowest $AIC$ value indicates the best model because that model strikes a better balance between fitting the data well and using fewer parameters (not over-fitting).
```{r, echo=FALSE}

models.aic |>
  dplyr::select( - Model2,
          - model) |>
  arrange( AIC ) |>
  mutate( P = format( P, scientific=TRUE, digits=3)) |>
  kable( caption = "<b>Which Variables Best Influence Male to Female Ratio?<b>",
         digits = 3) |>
   kable_classic( full_width = FALSE,
                 lightable_options = "striped") |>
  row_spec( 0, bold = TRUE)

```
Table 2. Male to Female Ratio Explained by Habitat Suitability, Phenotypes, and a combination of all variables. 

Q5: When plotting historical and present day habitat suitability, it is clear that there are in fact changes between the two time periods (Figure 5). However, it is slightly difficult to distinguish where the greatest change is occurring on this map overlay.   
```{r, echo=FALSE}

#read in raster
r2 <- raster("suitability_lgm.asc")

r2.utm <- projectRaster(r2, crs = new.proj)

#extract data
extract( r2.utm, sites) -> habitat_score2

sites$HabitatSuitHistorical <- habitat_score2

#change data format for easier visualization
sites_long <- pivot_longer(sites, 
                          cols = c("HabitatSuitHistorical", "HabitatSuitPresent"),
                          names_to = "Time_Period",
                          values_to = "Habitat_Suitability")

```

```{r, echo=FALSE}
#merge habitat scores w MF ratio values
merge( sites, Q1, by = "Site", all = FALSE) -> Q6

Q6 |>
  arrange( SiteNum ) |>
  dplyr::select( - Males,
          - Females)-> Q6

model.Q6 <- lm(MFRatio ~ HabitatSuitPresent + HabitatSuitHistorical, data = Q6)

Q6$Predicted_MFRatio_Historical <- predict(model.Q6, newdata = Q6)

Q6|>
  mutate( ΔHabitat = HabitatSuitPresent - HabitatSuitHistorical,
          ΔMFRatio = MFRatio - Predicted_MFRatio_Historical) -> Q6

```

```{r, echo=FALSE}

map_data("world") |>
  filter( region == "Mexico") -> map

min_value.P <- min(Q6$HabitatSuitPresent, na.rm = TRUE)
max_value.P <- max(Q6$HabitatSuitPresent, na.rm = TRUE)
min_value.H <- min(Q6$HabitatSuitHistorical, na.rm = TRUE)
max_value.H <- max(Q6$HabitatSuitHistorical, na.rm = TRUE)

ggplot(Q6) +
  geom_polygon( aes( x=long, 
                     y=lat, 
                     group=group ), 
                data=map, 
                fill="grey" ) +
  geom_sf(aes(color = HabitatSuitPresent), size = 3) +
  scale_color_gradient2(midpoint = 0, low = "blue", high = "red", mid = "white",
                        limits = c(min_value.P, max_value.H)) +
  labs(title = "Present Habitat Suitability",
       color = "Suitability",
       x = "Latitude",
       y = "Longitude") +
 theme_bw( base_size = 12 )  +
  theme(legend.position = "right")  +
  coord_sf( xlim = c(-115, -106),
            ylim = c(22, 30) ) +
  geom_sf_text(aes(label = SiteNum), size = 2.5, color = "black") -> Q5.map.P

```

```{r, echo=FALSE, warning=FALSE}

ggplot(Q6) +
  geom_polygon( aes( x=long, 
                     y=lat, 
                     group=group ), 
                data=map, 
                fill="grey" ) +
  geom_sf(aes(color = HabitatSuitHistorical), size = 3) +
  scale_color_gradient2(midpoint = 0, low = "blue", high = "red", mid = "white",
                        limits = c(min_value.P, max_value.H)) +  
  labs(title = "Historical Habitat Suitability",
       color = "Suitability",
       x = "Latitude",
       y = "Longitude") +
 theme_bw( base_size = 12 )  +
  theme(legend.position = "right")  +
  coord_sf( xlim = c(-115, -106),
            ylim = c(22, 30) ) +
  geom_sf_text(aes(label = SiteNum), size = 2.5, color = "black") -> Q5.map.H
```

```{r, echo=FALSE, warning=FALSE}
par(mfrow = c(1,2))
plot(Q5.map.P)
plot(Q5.map.H)
par(mfrow = c(1,1))
```
Due to this, another `ggplot`, this time a `geom_bar`, was used to take a closer look at each site. 19 of the 31 sites saw a decrease in habitat suitability, while only 12 saw an increase, indicating a majority trend for a decrease in habitat suitability from 20,000 years ago. It was also apparent that there was a larger change in habitat for the decreasing sites compared to the increasing sites (Figure 6).
```{r, echo=FALSE}

ggplot(Q6, aes(x = factor(SiteNum), y = ΔHabitat, fill = ΔHabitat > 0)) +  
  geom_bar(stat = "identity", color = "black") +  # Create bars with black border
  scale_fill_manual(values = c("red", "green")) +  # Set red for negative and green for positive
  labs(
    title = "Change in Habitat Suitability by Site", 
    x = "Site", 
    y = "Change in Habitat"
  ) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Figure 6. The change in habitat suitability from site to site. Red is a negative trend and green is a positive one.

```{r, echo=FALSE, message=FALSE}

#S-W normaility test
shapiro.test( Q6$ΔMFRatio) -> Shap.Ratio
shapiro.test( Q6$ΔHabitat) -> Shap.Hab

```

```{r, echo=FALSE}

#cor test for non normal data
cor.test( Q6$ΔMFRatio, Q6$ΔHabitat, method = "spearman" ) -> Q6.cor.test


```

Q6: The data tests show that the residuals of the `QQnorm` are normal, and the distribution of the newly created `Predicted_MFRatio_Historical` is also normal (Figure 7).      
```{r, echo=FALSE}

par(mfrow = c(1,2))
plot(model.Q6, which = 2, main = "Model.Q6 Q-Q")
hist(Q6$Predicted_MFRatio_Historical, main = "Predicted Histroical MFRatio Q-Q")
par(mfrow = c(1,1))

```
Figure 7. The `QQnorm` contains the Model.Q6 which shows the relationship `MFRatio ~ HabitatSuitPresent + HabitatSuitHistorical`. While the `histogram` shows the distribution of the `Predicted Histroical MFRatio`. 

Next, the variables `ΔHabitat` and `ΔMFRatio` were displayed visually to see if there were any noticeable trends or relationships between the two variables (Figures 8+9).  
```{r, echo=FALSE}

ggplot(Q6) +
  geom_polygon( aes( x=long, 
                     y=lat, 
                     group=group ), 
                data=map, 
                fill="grey" ) +
  geom_sf(aes(color = ΔMFRatio), size = 3) +  # Color by MFR Difference
  scale_color_gradient2(midpoint = 0, low = "blue", high = "red", mid = "white") +  # Blue for decrease, red for increase
  labs(title = "Differences in Male to Female Ratio: Historical vs. Present",
       color = "MFR Difference",
       x = "Latitude",
       y = "Longitude") +
 theme_bw( base_size = 12 )  +
  theme(legend.position = "right")  +
  coord_sf( xlim = c(-115, -106),
            ylim = c(22, 30) )

```
Figure 8. Differences in Male to Female Ratio: Historical vs. Present.
```{r, echo=FALSE}

ggplot(Q6) +
  geom_polygon( aes( x=long, 
                     y=lat, 
                     group=group ), 
                data=map, 
                fill="grey" ) +
  geom_sf(aes(color = ΔHabitat), size = 3) +  # Color by MFR Difference
  scale_color_gradient2(midpoint = 0, low = "blue", high = "red", mid = "white") +  # Blue for decrease, red for increase
  labs(title = "Differences in Habitat Suitability: Historical vs. Present",
       color = "Habitat Difference",
       x = "Latitude",
       y = "Longitude") +
 theme_bw( base_size = 12 )  +
  theme(legend.position = "right")  +
  coord_sf( xlim = c(-115, -106),
            ylim = c(22, 30) )

```
Figure 9. Differences in Male to Female Ratio: Historical vs. Present.

There were not any major trends noticed only visually, therefore statistical tests were ran to further confirm or deny this. In determining which test to run, it was found that the data behavior of `ΔHabitat` was not normal, but the `ΔMFRatio` variable was normal. Both a `Shapiro-Wilkes` and a `QQnorm` were ran on both variables. The `Shapiro-Wilkes` returned a value of $p$ = `r round(Shap.Ratio$p.value, 4)` for `ΔMFRatio`, and $p$ = `r round(Shap.Hab$p.value, 4)` for `ΔHabitat`, with a significant $p$-value of this test indicating non-normal distribution. The `QQnorm` plots also supported the `Shapiro-Wilkes` results (Figure 10). 
```{r, echo=FALSE}

#Visual normality test
par(mfrow = c(1,2))
qqnorm(Q6$ΔHabitat, main = "ΔHabitat")
qqline(Q6$ΔHabitat, col = "red")
hist(Q6$ΔHabitat, main = "ΔHabitat")
qqnorm(Q6$ΔMFRatio, main = "ΔMFRatio")
qqline(Q6$ΔMFRatio, col = "red")
hist(Q6$ΔMFRatio, main = "ΔMFRatio")
par(mfrow = c(1,1))

```
Figure 10. Distribution of data and residuals for `ΔHabitat` and `ΔMFRatio`.

The behavior of the data resulted in a `Spearman's Rho` correlation test being ran, which resulted in a $p$ = `r round(Q6.cor.test$p.value, 4)`. This obviously is not significant, meaning there does not seem to be significant relationship between the differences in climate (habitat suitability) and the male to female ratio.

## Appendix

```{r, echo=FALSE}

ggplot( Q3, aes( HabitatSuitPresent,
                      MFRatio)) +
      labs ( title = "Relationship Between Habitat Suitability and the Ratio of Male to Female Beetles",
             x = "Habitat Suitability",
             y = "Ratio of Male to Female") +
    stat_smooth( formula = y ~ x,
               method = "lm",
               alpha = 1) +
  geom_point( col = "red")

```
Figure 11. Correlation graph from question 3 showing the relationship between habitat suitability and male to female ratio.

```{r, echo=FALSE}

ggplot(sites_long, aes(x = Time_Period, y = Habitat_Suitability)) +
  geom_boxplot() +
  labs(title = "Distribution of Habitat Suitability by Time Period",
       x = "Time Period", y = "Habitat Suitability")


```
Figure 12. Scatter plot showing the distributions of habitat suitability historically and presently.

```{r, echo=FALSE}

ggplot(Q6, aes(x = ΔHabitat, y = ΔMFRatio )) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +  # Add regression line
  labs(title = "Change in MFR vs. Change in Habitat Suitability",
       x = "Change in Habitat Suitability",
       y = "Change in MFR")


```
Figure 13. Scatter plot showing the relationship between the ΔHabitat and the ΔMFRatio.
