---
title: "Rice Rivers Center Tree Inventory"
author: "Eric Bouma"
format: html
  #html:             #to turn code off globally
    #echo: false
    #warning: false
    #message: false
    #error: false


---

## **Abstract**

The Rice Rivers Center (RRC) Tree Inventory is a data set focused on tree species within the property of the center. This document contains the RRC Tree Inventory data set, along with code for data wrangling involving summary statistics for a few columns of data. The goal of this document is to highlight a few patterns within the RRC Tree Inventory.

```{r, include = FALSE, echo = FALSE}
#if want to only disable chunk
# #| message: false
# #| warning: false
# #| echo: false


#loading packages
library(tidyverse)
library( lubridate )
library( dplyr)
library( knitr )
library( kableExtra )

```

```{r, include = FALSE, echo = FALSE}
#Reading in the data

read.csv( "TreeInv_DEM_extract_ExportTable.csv" ) -> rrc_ti
```

## **Introduction**

Tree data was collected within a roughly 28 acre area of interest (AOI) at the RRC. Tree data includes the common and scientific names, diameter at breast height (DBH), four letter ID code, and a unique alpha-numeric ID code for each tree. In addition to the tree data, this data set contains carbon and elevation metrics for each tree. Data types include numerical, categorical, and character.

```{r, include = FALSE, echo = FALSE}
#Looking at the structure of the data
str( rrc_ti )
```

## **Methods**

The methods for summarizing statistics involved four main comparisons; the numerical distribution trees species, amount of carbon stored by species, DBH by species, and DBH by elevation class. The functions mutate, count, summarize, group_by, and arrange were all used.

**Distribution of Species**

```{r}
rrc_ti |> 
  mutate( `Common Name` = Common_name) |>
  count( `Common Name`,
         sort = TRUE,
         name = "Species Distribution")

```

**Amount of Carbon Stored by Species**

```{r}

rrc_ti |>
   mutate( `Common Name` = Common_name) |>
  group_by( `Common Name` ) |>
  summarize( `Avg Carbon Stored (Kg)` = mean( Carbon_Stored ),
             `Total Carbon Stored (Kg)` = sum( Carbon_Stored)) |>
  arrange( - `Avg Carbon Stored (Kg)`,
           - `Total Carbon Stored (Kg)`)
```

**DBH by Species**

```{r}

rrc_ti |>
  mutate( `Common Name` = Common_name) |>
  group_by( `Common Name`) |>
  summarize( `Avg DBH (cm)` = mean( DBH ),
             `Max DBH (cm)` = max( DBH )) |>
  arrange( - `Avg DBH (cm)`,
           - `Max DBH (cm)`)
```

**DBH by Elevation Type**

```{r}
rrc_ti |>
   mutate( `Elevation` = DEM_POINTS) |>
  group_by( Elevation ) |>
  summarize( `Avg DBH (cm)` = mean( DBH ),
             `Max DBH (cm)` = max( DBH )) |>
  arrange( - `Avg DBH (cm)`)
  
```

## **Results**

**Distribution of Species**

```{r}
rrc_ti |> 
  mutate( `Common Name` = Common_name) |>
  count( `Common Name`,
         sort = TRUE,
         name = "Species Distribution") -> table1 
  
table1 |>
  kable( digits = 0,
         caption = "Number of Trees per Species") -> kbl1

kbl1 |>
   kable_classic( full_width = FALSE,
                 lightable_options = "striped") |>
  row_spec( 0, bold = TRUE)

```

**Amount of Carbon Stored by Species**

```{r}

rrc_ti |>
  mutate( `Common Name` = Common_name) |>
  group_by( `Common Name` ) |>
  summarize( `Avg Carbon Stored (Kg)` = mean( Carbon_Stored ),
             `Total Carbon Stored (Kg)` = sum( Carbon_Stored)) |>
  arrange( - `Avg Carbon Stored (Kg)`,
           - `Total Carbon Stored (Kg)`) -> table2

table2 |>
  kable( digits = 2,
         caption = "Kg of Carbon Stored by Species") -> kbl2

kbl2 |>
  kable_classic( full_width = FALSE,
                 lightable_options = "striped") |>
  row_spec( 0, bold = TRUE)
```

**DBH by Species**

```{r}

rrc_ti |>
  mutate( `Common Name` = Common_name) |>
  group_by( `Common Name`) |>
  summarize( `Avg DBH (cm)` = mean( DBH ),
             `Max DBH (cm)` = max( DBH )) |>
  arrange( - `Avg DBH (cm)`,
           - `Max DBH (cm)`) -> table3

table3 |>
  kable( digits = 2,
         caption = "DBH by Species") -> kbl3

kbl3 |>
   kable_classic( full_width = FALSE,
                 lightable_options = "striped") |>
  row_spec( 0, bold = TRUE)
```

**DBH by Elevation Type**

```{r}

rrc_ti |>
  mutate( `Elevation` = DEM_POINTS) |>
  group_by( Elevation ) |>
  summarize( `Avg DBH (cm)` = mean( DBH ),
             `Max DBH (cm)` = max( DBH )) |>
  arrange( - `Avg DBH (cm)`) -> table4

table4 |>
  kable( digits = 2,
         caption = "DBH by Elevation") -> kbl4

kbl4 |>
  kable_classic( full_width = FALSE,
                 lightable_options = "striped") |>
  row_spec( 0, bold = TRUE)
```
