---
title: "**Textual Data HW**"
author: "Eric Bouma"
format: html
editor: visual
---

## **Abstract**

The purpose of this assignment is to analyze the course titles from all undergraduate offerings in VCU Life Sciences using textual data functions (stringr) . The data for this assignment is a text (.txt) file, with a textual "dump" of all course descriptions from Bioinformatics (BNFO), Environmental Studies (ENVS), and Life Sciences (LFSC).

```{r, message=FALSE, echo=FALSE}

#Read Packages

library(tidyverse)
library(readr)
library(stringr)
library( wordcloud)
library(tm)

```



```{r, echo=FALSE}
#Read in Data

lfsc <- read_lines("LifeSciencesCurriculum.txt")

#head(lfsc)

```
## **List of All Courses**
```{r}

str_detect(lfsc, "BNFO|ENVS|LFSC") -> idx1

lfsc [idx1]

```
## **Pulling Out Course Info**
```{r}

#str_view(lfsc, "^[A-Z]{4} [0-9]{3}.+ [0-9] Hour[s]?\\.$")

pattern <- "^[A-Z]{4} [0-9]{3}.+ [0-9] Hour[s]?\\.$"

grepl( pattern, lfsc) -> idx1

courses <- lfsc[idx1]
#courses

```
## **Taking Selected Data and Putting it into a Data Frame**
```{r}

str_split( courses, pattern = "\\.",
           simplify = TRUE) -> table1
#table1

data.frame ( Course = table1[,1],
                     Title = str_trim(table1[,2]),
                     Hours = str_trim(table1[,3])) |>
  mutate( Hours = str_remove( Hours, " Hour")) |>
  mutate( Hours = str_remove( Hours, "s"))  |>
  mutate( Program = str_split( Course, " ", simplify = TRUE)[,1]) |>
  mutate( Number = as.numeric (str_split( Course, " ", simplify = TRUE)[,2])) |>
  select( Program, Number, Title, Hours) -> data
```

## **Bar Plot of Courses by Level Grouped by Program**
```{r}

library(RColorBrewer)

data |>
  mutate( `Course Level` = case_when(
    Number >= 100 & Number <= 199 ~ 100,
    Number >= 200 & Number <= 299 ~ 200,
    Number >= 300 & Number <= 399 ~ 300,
    Number >= 400 & Number <= 499 ~ 400)) -> data

data |>
  ggplot( aes( `Course Level`,
               fill = Program)) +
  labs ( title = "Number of Courses by Level Grouped by Program",
         x = "Course Level",
         y = "Count") +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  theme_update()

```

## **Word Cloud**
```{r, message=FALSE, warning=FALSE}

wordcloud( words = data$Title,
           min.freq = 1,
           scale = c(3,0.5),
           colors = brewer.pal(8, "Dark2"),
           random.order = F,
           rot.per = 0.25)


```

